---
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
---

```{=html}
<style>
  code {
    font-size: 1em;
  }
  p {
    text-align: justify;
  }
</style>
```

# Visual Analytics for AI Multi-Agent System Diagnostics

## 1. Overview

Modern enterprises increasingly deploy AI-powered multi-agent systems to automate routine workflows — drafting communications, managing files, routing tasks, and maintaining social media presence. These systems offer significant efficiency gains, but introduce a new class of operational risk - when agents malfunction or interact in unintended ways, the consequences can propagate silently across systems before any human notices.

This study investigates such an incident. On 17 May 2046 at 4:21am local time, an AI agent associated with John Windward — an employee of TenantThread (Company A) — published an anomalous post on SaidIT, a public social media platform. The post was routed to a seemingly random forum and contained content that appeared to be gibberish rather than intentional communication. No corresponding user action triggered the post.

Visual analytics — the integration of data visualisation, statistical reasoning, and interactive exploration — provides the investigative framework for this study. By constructing timelines, network maps, and event-sequence comparisons from TenantThread's internal system logs, we move beyond simple log inspection to reveal the causal chain, contextualise the incident within the broader system architecture, and propose a targeted intervention to prevent recurrence.

### 1.1 Objectives

This study applies visual analytics techniques in R to TenantThread's multi-agent event log, spanning 8 May to 16 Jul 2046.

The specific objectives are:

**1. Reconstruct the Incident**: (a) Identify the detailed chain of events that led to the anomalous SaidIT post, and (b) map the system architecture within which it occurred.

**2. Decode the Content**: Determine what the anomalous post contains and trace the origin of its content within TenantThread's file and agent activity logs.

**3. Prevent Recurrence**: (a) Identify whether similar anomalous behaviour has occurred previously in the log, and (b) propose a single, well-defined intervention point to prevent future occurrences.

## 2. The Data

The dataset used in this study is drawn from the internal event logs of TenantThread's multi-agent AI system, made available as part of [VAST Challenge 2026 Mini-Challenge 2](https://vast-challenge.github.io/2026/MC2.html). It captures a continuous record of all automated and human-initiated actions within the system from 8 May to 16 Jul 2046.

There are two files provided:

- **mc2_df.rds**: The primary event logs in a 185,147 rows R dataframe. Each row represents a single system event. The data spans 24 distinct event types ranging from routine file operations and email checks to social media posts and agent-to-agent task assignments.

- **org_chart.json**: This file described TenantThread's organisational hierarchy, containing 75 nodes (1 company, 6 departments, 19 teams, and 49 persons) connected by hierarchical "contains" relationships.

Together, these two files allow us to reconstruct both the *what happened* (event log) and the *who is involved* (org chart) dimensions of the anomalous posting incident.

## 3. Getting Started

### 3.1 Loading R Packages

The following R packages will be loaded using `pacman::p_load()`, which automatically installs any missing packages before loading them.

- **tidyverse**: Core collection of R packages for data wrangling and visualisation, including ggplot2, dplyr, tidyr, readr, and lubridate.
- **jsonlite**: Parses JSON files into R objects. Used to load the organisational hierarchy from `org_chart.json`.
- **lubridate**: Simplifies working with Unix timestamps — converting them to datetime objects and extracting components such as date, hour, and weekday.
- **here**: Constructs file paths relative to the project root, ensuring code runs consistently regardless of the working directory.
- **scales**: Formats axis labels and annotations with commas, percentages, and readable number formats.
- **ggtext**: Enables rich text formatting (bold, colour, italics) within ggplot2 titles and labels using HTML or Markdown syntax.
- **patchwork**: Combines multiple ggplot2 plots into a single composited layout with shared titles and captions.
- **plotly**: Converts ggplot2 charts into interactive HTML visualisations with hover tooltips and zoom controls.
- **visNetwork**: Renders interactive network graphs in HTML, allowing nodes and edges to be explored through drag, zoom, and click interactions.
- **tidygraph**: Provides a tidy interface for graph manipulation, enabling `dplyr`-style operations on network nodes and edges.
- **ggraph**: Extends ggplot2 with geoms for static network visualisation, built on top of `tidygraph`.
- **igraph**: Provides graph computation algorithms — used for centrality measures and shortest-path calculations on the org chart and event networks.
- **DT**: Renders R dataframes as interactive HTML tables with sorting, filtering, and pagination.
- **knitr**: Formats R dataframes as clean, publication-ready tables within the HTML output using `kable()`.
- **ggrepel**: Extends ggplot2 by automatically repositioning overlapping text and label annotations, drawing connecting segments back to their original data points to preserve readability.

```{r}
#| label: packages
#| code-fold: true
#| code-summary: "Show code"
#| warning: false
pacman::p_load(tidyverse, jsonlite, lubridate, here, scales, ggtext, patchwork, plotly, visNetwork, tidygraph, ggraph, igraph, DT, knitr, ggrepel)
```

### 3.2 Importing the Data

```{r}
#| code-fold: true
mc2_df <- readRDS("data/raw/mc2_df.rds")
org_chart <- fromJSON("data/raw/org_chart.json")

cat("mc2_df rows:", nrow(mc2_df), "\n")
cat("mc2_df cols:", ncol(mc2_df), "\n")
cat("Date range:",
    format(as_datetime(min(mc2_df$when)), "%Y-%m-%d"), "to",
    format(as_datetime(max(mc2_df$when)), "%Y-%m-%d"), "\n")
```

### 3.3 Data Preparation

Two transformations are applied before any analysis begins.

- The `when` column, stored as a unix timestamp (seconds since 1 January 1970), is converted to a proper datetime object using `as_datetime()`. Three derived time columns are added: `date` (calendar date), `hour` (hour of day in UTC), and `wday` (day of week). These are used in the activity heatmap and event timeline visualisations in subsequent sections.

- The `details` column contains a nested dataframe with 42 sub-columns embedded within each row of `mc2_df`. These sub-columns hold the event-specific attributes critical to this investigation, including `forum`, `content`, `content_source`, `file`, `from`, `to`, and `virus`. To enable direct column-level access without repeated `$details$` indexing, the nested dataframe is flattened into the main dataframe using `bind_cols()`.

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| include: false

mc2_df <- mc2_df |>
  mutate(
    datetime = as_datetime(when),
    date     = as_date(datetime),
    hour     = hour(datetime),
    wday     = wday(datetime, label = TRUE, abbr = TRUE)
  )

details_flat <- mc2_df$details
mc2_df       <- mc2_df |> bind_cols(details_flat)
```

## 4. Data Understanding

Before investigating the anomalous post, we first orient ourselves to the full dataset — what types of events were logged, how frequently each occurs, and when the system was most active. This baseline understanding is essential for identifying what is *normal* system behaviour before contrasting it against the specific anomaly of interest.

### 4.1 Event Type Overview

An interactive table is used to display the frequency of all 24 event types recorded in the dataset. Sorting by count allows me to immediately identify which automated behaviours dominate the system and which — such as `post_saidit` and `saidit_post` — are relatively rare. This rarity is analytically significant: low-frequency event types are more likely to contain anomalies that would be invisible in aggregate reporting.

### 4.2 Event Frequency Table

```{r}
#| label: event-frequency
#| code-fold: true
#| code-summary: "Show code"

event_counts <- mc2_df |>
  count(short_name, sort = TRUE) |>
  rename(`Event Type` = short_name,
         Count = n) |>
  mutate(`% of Total` = percent(Count / sum(Count), 
                                accuracy = 0.1))

DT::datatable(
  event_counts,
  options  = list(pageLength = 24, 
                  dom        = "t",
                  order      = list(list(1, "desc"))),
  rownames = FALSE,
  caption  = "Table 4.1: Event type frequencies (n = 185,147)"
)
```

::: callout-note
## Observations

- The event log is dominated by routine automated behaviours, like `check_email` along accounts for 37,286 events (20% of all activity), followed by `assign_agent_task` at 18,856 and `read_file` at 18,838. Together, the top five event types account for over 57% of all logged events, confirming that the vast majority of system activity is mundane and repetitive.

- The social media posting events are notably rare by comparison: `post_saidit` appears 133 times, `saidit_post` 108 times, and `saidit_post_check` 71 times (collectively less than 0.2% of all events). This rarity means any deviation in their structure or sequencing stands out sharply against the backgroun noise of the event log, and motivates the focused investigation following this.
:::

### 4.3 System Activity Heatmap

A day-of-week by hour-of-day heatmap is used to reveal the temporal rhythm of system activity across the full observation period. This visualisation is preferred over a simple daily bar chart because it simultaneously captures two temporal dimensions: *when in the week* and *when in the day* - making it possible to identify both business-hours patterns and off-hours anomalies in a single view.

### 4.4 Activity Heatmap

```{r}
#| label: activity-heatmap
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Figure 4.1: System activity heatmap by weekday and hour (UTC)"

mc2_df |>
  count(wday, hour) |>
  ggplot(aes(x = hour, y = wday, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low  = "#f7fbff",
    high = "#08519c",
    name = "Event count"
  ) +
  scale_x_continuous(
    breaks = seq(0, 23, by = 2),
    labels = paste0(seq(0, 23, by = 2), ":00")
  ) +
  labs(
    title    = "When Is the System Most Active?",
    subtitle = "Heatmap of event counts by weekday and hour (UTC)",
    x        = "Hour of Day (UTC)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid  = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title  = element_text(face = "bold")
  )
```

::: callout-note
## Observations

- The heatmap reveals a distinctly uneven distribution of system activity across the week. The single busiest period is **Saturday between 08:00 and 12:00 UTC**, with event density peaking sharply around 09:00 to 10:00 - suggesting a scheduled batch of automated agent workflows concentrated on Saturday mornings. Firday shows two separate activity bands: an early morning cluster (00:00 to 02:00 UTC) and an evening cluster (16:00 to 22:00 UTC), while Thursday activity is concentrated in the late evening (20:00 to 23:00 UTC). Mid-week days are largely quiet.

- The temporal baseline is directly relevant to the anomalous post to be investigated next. The post was recorded at **11:21 UTC on Sunday 17 May 2046** and Sunday is the quietest day in the entire observation period, with virtually no system activity recorded at that hour. The near-absence of surrounding automated activity at that time is consistent with a rogue process executing in isolation, without the oversight of concurrent agent workflows that could normally be present during peak hours.
:::

## 5. How was the Post Made?

To reconstruct how the anomalous SaidIT post came to exist on May 17, 2046, we begin by establishing what normal posting activity looks like in this system, and then systematically identify where the May 17 event deviates from it.

The event log contains three posting-related event types that are directly relevant to this investigation: - **`post_saidit`** - the instruction event, where a human or agent requests that a post be made on SaidIT. - **`saidit_post`** - the confirmation event, where the system logs that a post was actually published on the forum. This event contains the `content` and `content_source` fields that tell us *what* was posted and *where it came from*. - **`saidit_post_check`** - a verification event that confirms whether the post was successfully made.

We focus our analysis on **`saidit_post`** because it is the only event that carries direct evidence of what was published — specifically the `content` and `content_source` fields. An anomaly in `saidit_post` means something abnormal was actually posted, not merely attempted. This makes it the most forensically relevant event type.

### 5.1 Normal vs Anomalous Posting Pattern

To establish a baseline, we classify all `saidit_post` events into normal and anomalous patterns based on whether `content_source` is populated. Under normal operation, an agent generates post content directly - `content` is populated with text and `content_source` remains null, indicating no external file was involved. An anomalous post reverses this - its `content` will be null and `content_source` points to an external file, indicating the post content was injected from outside the agent workflow.

```{r}
#| label: posting-pattern-comparison
#| code-fold: true
#| code-summary: "Show code"

saidit_events <- mc2_df %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    pattern = case_when(
      !is.na(content_source) ~ "Anomalous",
      TRUE                   ~ "Normal"
    )
  )

pattern_summary <- saidit_events %>%
  group_by(pattern) %>%
  summarise(
    Count            = n(),
    `content field`  = ifelse(all(is.na(content)),  "NULL",     "Text present"),
    `Files involved` = paste(unique(na.omit(content_source)), collapse = ", "),
    `content_source` = ifelse(all(is.na(content_source)), "NULL", "Filename (.txt)"),
    .groups = "drop"
  )

kable(pattern_summary,
      caption = "Table 5.1: Normal vs Anomalous saidit_post Events")
```

::: callout-note
## Observations

- Of the **108 total `saidit_post` events** in the dataset, **105 follow the normal pattern** — `content` is populated with agent-generated text and `content_source` is null.

- **3 posts are anomalous**: their `content` field is null and `content_source` points to an external `.txt` file — **SwiftWren.txt**, **HiddenOrca.txt**, and **MellowOtter.txt**. It confirms that the anomalous posts were not generated by the agent, but its content was **injected directly from an external file**, bypassing normal agent-generated content entirely.

- This baseline distinction between normal and anomalous posting behaviour raises an immediate question: which agent or system is responsible for all three anomalous posts? This is examined in Section 5.2.
:::

### 5.2 Identifying the Agent Behind the Anomalous Posts

First we establish *who* is the focus of the investigation. The three anomalous `saidit_post events` are filtered and their associated agent fields inspected to determine which person or system is linked to the injection.

```{r}
#| label: anomalous-files-to-agent
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(short_name == "saidit_post",
         !is.na(content_source)) %>%
  mutate(parties_str = map_chr(parties, ~ paste(.x, collapse = ", "))) %>%
  arrange(datetime) %>%
  select(datetime, content_source, parties_str) %>%
  rename(`Datetime (UTC)`    = datetime,
         `Injected File`     = content_source,
         `Agents in Parties` = parties_str) %>%
  kable(caption = "Table 5.2a: Agents linked to each anomalous saidit_post event")
```

```{r}
#| label: anomalous-post-detail
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    pattern     = if_else(is.na(content_source), "Normal", "Anomalous"),
    parties_str = map_chr(parties, ~ paste(.x, collapse = ", ")),
    involves_jw = str_detect(tolower(parties_str), "john_windward")
  ) %>%
  count(pattern, involves_jw, content_source) %>%
  mutate(
    involves_jw    = if_else(involves_jw, "Present", "Absent"),
    content_source = if_else(is.na(content_source), "—", content_source)
  ) %>%
  rename(
    Pattern          = pattern,
    `John Windward`  = involves_jw,
    `Injected File`  = content_source,
    Count            = n
  ) %>%
  arrange(Pattern, `John Windward`) %>%
  kable(caption = "Table 5.2b: saidit_post events by pattern, injected file, and John Windward's presence")
```

::: callout-note
## Observations

- **Table 5.2a** shows that `Agent/person:john_windward` appears in the `parties` field of every anomalous `saidit_post` event — across all three injected files **HiddenOrca.txt** (10 May), **MellowOtter.txt** (11 May), and **SwiftWren.txt** (17 May). No other human agent is consistently present across all three events, making John Windward the **unique common thread** linking every file-injected post in the dataset.

- **Table 5.2b** contextualises this finding: of John Windward's **22 total `saidit_post` events**, 19 follow the normal pattern and only 3 are anomalous. His regular posting activity confirms he is a **legitimate system participant** whose agent was exploited — his presence alone does not flag a post as anomalous. What is significant is that **all 3 anomalous posts involve him, without exception**, directly motivating the focused timeline investigation in Section 5.3.
:::

### 5.3 May 17 Event Timeline - John Windward

A chronological timeline of all events involving John Windward on May 17, 2046 provides essential context for the anomalous post at 11:21 UTC. A dot-and-line timeline plot is used rather than a table, because it makes the *temporal gaps and clustering* between events immediately visible. The anomalous `saidit_post` is highlighted in red to anchor the pivotal event within its surrounding activity sequence.

```{r}
#| label: may17-timeline-twopanel
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 11
#| fig-height: 8

jw_may17 <- mc2_df %>%
  filter(
    date == as.Date("2046-05-17"),
    map_lgl(parties, ~any(str_detect(.x, "john_windward")))
  ) %>%
  mutate(
    is_anomaly = short_name == "saidit_post" & !is.na(content_source)
  )

# ── Panel 1: Hourly event count bar chart ────────────────────────────────────

p_hourly <- jw_may17 %>%
  mutate(hour = lubridate::floor_date(datetime, "hour")) %>%
  count(hour) %>%
  mutate(is_anomaly_hour = (hour == lubridate::floor_date(
    as.POSIXct("2046-05-17 11:21:00", tz = "UTC"), "hour"))) %>%
  ggplot(aes(x = hour, y = n, fill = is_anomaly_hour)) +
  geom_col(width = 2400) +
  geom_vline(xintercept = as.POSIXct("2046-05-17 11:21:00", tz = "UTC"),
             colour = "#E15759", linewidth = 0.8, linetype = "dashed") +
  annotate("text",
           x     = as.POSIXct("2046-05-17 11:21:00", tz = "UTC"),
           y     = Inf, vjust = 1.5, hjust = -0.1,
           label = "Anomalous post\n11:21 UTC",
           colour = "#E15759", size = 3.2, fontface = "bold") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "#E15759"),
                    guide = "none") +
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "2 hours") +
  labs(title = "Hourly Event Count — John Windward, May 17 2046",
       x = "Hour (UTC)", y = "Event Count") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# ── Panel 2: Zoomed window 10:30–12:00 ───────────────────────────────────────

zoom_start <- as.POSIXct("2046-05-17 10:30:00", tz = "UTC")
zoom_end   <- as.POSIXct("2046-05-17 12:00:00", tz = "UTC")

p_zoom <- jw_may17 %>%
  filter(datetime >= zoom_start, datetime <= zoom_end) %>%
  ggplot(aes(x = datetime, y = 1)) +
  geom_line(colour = "grey70", linewidth = 0.5) +
  geom_point(aes(colour = is_anomaly, size = is_anomaly)) +
  geom_text_repel(
    data = jw_may17 %>%
      filter(datetime >= zoom_start, datetime <= zoom_end, !is_anomaly),
    aes(label = short_name),
    size          = 3.2,
    colour        = "grey40",
    nudge_y       = 0.3,
    direction     = "x",
    segment.color = "grey70",
    segment.size  = 0.3,
    max.overlaps  = Inf
  ) +
  geom_label_repel(
    data = jw_may17 %>%
      filter(datetime >= zoom_start, datetime <= zoom_end, is_anomaly),
    aes(label = paste0(short_name, "\n[SwiftWren.txt]")),
    size          = 3.2,
    colour        = "#E15759",
    fill          = "#FFF0F0",
    fontface      = "bold",
    label.size    = 0.3,
    nudge_y       = -0.4,
    direction     = "x",
    segment.color = "#E15759",
    segment.size  = 0.4
  ) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "#E15759"),
                      guide = "none") +
  scale_size_manual(values = c("FALSE" = 2.5, "TRUE" = 5),
                    guide = "none") +
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "10 mins",
                   limits = c(zoom_start, zoom_end)) +
  labs(title = "Zoomed View: 10:30–12:00 UTC",
       x = "Time (UTC)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title         = element_text(face = "bold")
  )

# ── Combined layout ───────────────────────────────────────────────────────────

p_hourly / p_zoom +
  plot_annotation(
    title   = "Figure 5.3: John Windward Event Timeline — May 17, 2046",
    caption = "Source: mc2_df | Dashed red line marks anomalous saidit_post at 11:21 UTC",
    theme   = theme(
      plot.title   = element_text(face = "bold", size = 13),
      plot.caption = element_text(size = 9, colour = "grey50")
    )
  )
```

::: callout-note
## Observations

- John Windward's activity on May 17 is **concentrated in four temporal clusters**: a burst of 7 events at **07:00 UTC**, 5 events around **11:00 UTC**, a single `read_file` at **17:00 UTC**, and the largest cluster of 8 events at **23:00 UTC**.

- The anomalous `saidit_post` at **11:21 UTC** occurs alongside four other events - `queue_subordinate_task`, `saidit_post_check`, and two `delete_file` events. All these events happened within the same minute, suggesting a **coordinated sequence rather than an isolated action**.

- Critically, the two `delete_file` events occuring immediately alongside the anomalous post suggest a possible **cover up attempt** - files deleted to remove evidence of how the post was generated.

- The `read_file` event at **17:00 UTC**, hours after the incident, warrants further investigation into what file was accessed and whether it is connected to the SwiftWren.txt injection.
:::

::: callout-warning
## Caveat

The `parties` field is a list column, meaning events are filtered by detecting the string `"john_windward"` within each list entry. Events where Windward participated as a background system participant rather than a named party may not appear in this timeline.
:::

## 6. System Overview

Having established *how* the anomalous post was created - injected from an external `.txt` file via John Windward's agent - and *who* is linked to it, the investigation now turns to *where* John Windward sits within TenantThread's organisation structure and how his agent came to execute the anomalous post.

This section is organised into three parts. Section 6.1 maps the full organisational hierarchy to contextualise John Windward's position within TenantThread. Section 6.2 traces the `queue_subordinate_task` events that converged at the moment of the anomalous post to identify whether John Windward acted alone or was instructed by another agent. Section 6.3 then constructs his communication network in the 48hrs window surrounding May 17 to surface the full pattern of interactions leading up to and following the incident.

### 6.1 Organisational Hierarchy

To contextualise the positions of all persons relevant to this investigation, the full 75-node TenantThread organisational chart is rendered as an interactive network using `visNetwork`. This package is chosen over static alternatives such as `ggraph` because the sheer number of nodes - spanning one company, six departments, 19 teams, and 49 persons - would render a fixed-size plot illegible. The interactive canvas allows us to zoom, pan, and hover on individual nodes to inspect relationships without visual compression. A top-down hierarchical layout is applied using `visHierarchicalLayout()` to preserve the reporting structure inherent in the `contains` edge type. **John Windward** is highlighted in red to immediately locate his position within the hierarchy and establish the organisational contxt before investigating how his agent came to execute the anomalous post.

```{r}
#| code-fold: true
#| label: org-chart-network
#| code-summary: "Show code"
#| out-width: "100%"

library(visNetwork)

# ── Parse org chart JSON ──────────────────────────────────────────────────────
org_raw <- fromJSON("data/raw/org_chart.json", simplifyDataFrame = FALSE)

nodes_df <- map_dfr(org_raw$nodes, ~ tibble(
  id    = .x$id,
  label = .x$label,
  type  = .x$type
))

edges_df <- map_dfr(org_raw$edges, ~ tibble(
  from = .x$source,
  to   = .x$target
))

# ── Assign visual properties ──────────────────────────────────────────────────
nodes_vis <- nodes_df %>%
  mutate(
    color.background = case_when(
      type == "company"    ~ "#2C3E50",
      type == "department" ~ "#2980B9",
      type == "team"       ~ "#27AE60",
      type == "person"     ~ "#D5E8F3",
      TRUE                 ~ "#BDC3C7"
    ),
    color.border = "white",
    font.color = case_when(
      type %in% c("company", "department", "team") ~ "white",
      TRUE                                         ~ "#2C3E50"
    ),
    size = case_when(
      type == "company"    ~ 40,
      type == "department" ~ 28,
      type == "team"       ~ 18,
      type == "person"     ~ 14,
      TRUE                 ~ 12
    ),
    shape = case_when(
      type == "company"    ~ "diamond",
      type == "department" ~ "box",
      type == "team"       ~ "box",
      TRUE                 ~ "dot"
    ),
    # Highlight John Windward in red
    color.background = if_else(
      str_detect(tolower(id), "john_windward") |
        str_detect(tolower(label), "john windward"),
      "#E74C3C", color.background
    ),
    font.color = if_else(
      str_detect(tolower(id), "john_windward") |
        str_detect(tolower(label), "john windward"),
      "white", font.color
    ),
    size = if_else(
      str_detect(tolower(id), "john_windward") |
        str_detect(tolower(label), "john windward"),
      22, size
    ),
    title = paste0("<b>", label, "</b><br>Type: ", type)
  )

# ── Render ────────────────────────────────────────────────────────────────────
visNetwork(
  nodes_vis,
  edges_df,
  main   = list(
    text  = "Figure 6.1: TenantThread Organisational Hierarchy",
    style = "font-family:sans-serif; font-weight:bold; font-size:14px; text-align:center;"
  ),
  height = "620px",
  width  = "100%"
) %>%
  visHierarchicalLayout(
    direction            = "UD",
    levelSeparation      = 120,
    nodeSpacing          = 45,
    treeSpacing          = 150,
    blockShifting        = TRUE,
    edgeMinimization     = TRUE,
    parentCentralization = TRUE
  ) %>%
  visNodes(borderWidth = 1.5, borderWidthSelected = 3) %>%
  visEdges(
    arrows = list(to = list(enabled = TRUE, scaleFactor = 0.6)),
    color  = list(color = "#95A5A6", opacity = 0.6),
    smooth = list(type = "cubicBezier")
  ) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, main = "Select a node")
  ) %>%
  visInteraction(navigationButtons = TRUE, tooltipDelay = 100) %>%
  visLegend(
    position = "left",
    width    = 0.10,
    addNodes = list(
      list(label = "Company",       shape = "diamond", color = "#2C3E50",
           font  = list(color = "white")),
      list(label = "Department",    shape = "box",     color = "#2980B9",
           font  = list(color = "white")),
      list(label = "Team",          shape = "box",     color = "#27AE60",
           font  = list(color = "white")),
      list(label = "Person",        shape = "dot",     color = "#D5E8F3",
           font  = list(color = "#2C3E50")),
      list(label = "John Windward", shape = "dot",     color = "#E74C3C",
           font  = list(color = "white"))
    ),
    useGroups = FALSE
  ) %>%
  visEvents(type = "once", afterDrawing = "function() { this.fit(); }")
```

::: callout-note
## Observations — Organisational Position

- The TenantThread hierarchy comprises **75 nodes**: 1 company root, 6 departments, 19 teams, and 49 individual persons, all connected by directional `contains` edges flowing from parent to child.

- **John Windward** (highlighted in red) reports directly to the **Customer Support** department and is not assigned to any sub-team. He has zero outgoing `contains` edges, confirming he holds **no formal managerial authority** over any other person or system in the organisational chart.

- Despite having no formal subordinates, a `queue_subordinate_task` event appears in John Windward's activity at**11:21 UTC on May 17** — the exact moment of the anomalous post. Whether he issued or received this instruction, and who else was involved, will be examined in Section 6.2.
:::

### 6.2 Tracing the Task Chain

Section 6.1 established that John Windward was present in the `parties` field of all three anomalous posts and that the `queue_subordinate_task` event at **11:21 UTC on May 17** - the event timestamp of the third anomalous `saidit_post`. The immediate investigative question is whether John Windward *issued* that task or *received* it. The `from` and `to` fields are `NA` for most events in this dataset; directionality for task delegation events is instead encoded in the `target_agent` field, which records the agent designated to carry out the task. Re-running the original discovery query as a documented code chunk - filtered to all `queue_subordinate_task` events at 11:21 UTC on May 17, is the first step in resolving this question.

```{r}
#| code-fold: true
#| label: task-chain-discovery
#| code-summary: "Show code"

mc2_df %>%
  filter(short_name == "queue_subordinate_task",
         date == as.Date("2046-05-17"),
         hour == 11) %>%
  select(datetime, short_name, task, target_agent,
         parties) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = " → "))) %>%
  select(-parties) %>%
  arrange(datetime) %>%
  knitr::kable(
    col.names = c("Timestamp (UTC)", "Event Type", "Task",
                  "Target Agent", "Parties"),
    caption   = "Table 6.2: All queue_subordinate_task events — Hour 11, May 17 2046"
  )
```

::: callout-note
## Observations

- The query returns **two events**, not one. An earlier delegation at **11:12:24** preceded the 11:21:13 event identified in Section 6.1.
- In both events, `from` and `to` are `NA`. **Directionality is established entirely through `target_agent`**, which names the agent designated to receive and execute the task — and through `parties`, which lists the issuer and recipient in order.
- At **11:21:13**, `target_agent` is `Agent/person:john_windward`. This is the key proof: **John Windward did not issue the task — he received it.** The agent listed first in `parties` is the issuer; John appears second.
- At **11:12:24**, the same `read_file` task was issued to a different agent, who appears as the *issuer* in the 11:21:13 event nine minutes later.
- A **three-person relay chain** is confirmed, culminating at the timestamp of the anomalous post.
:::

With directionality established through `target_agent`, the full delegation chain can now be named. The first event shows **Daniel Gangway** tasking **Chloe Ballast** to `read_file` at 11:12. Chloe then relays the same instruction onward to **John Windward** at 11:21 — the moment the anomalous post is recorded. The chain is: **Daniel Gangway → Chloe Ballast → John Windward**.

A directed network diagram is used to make this relay immediately legible. A chain layout is preferred over a full communication network here because the finding is structurally simple — three nodes, two directed edges, two timestamps — and any additional nodes would dilute the causal signal. Edge labels carry the timestamps to anchor the visualisation to the evidence in Table 6.2.

```{r}
#| code-fold: true
#| label: task-chain-network
#| code-summary: "Show code"

chain_vis_nodes <- tibble(
  id        = 1:3,
  label     = c("Daniel Gangway\n(issuer)", 
                "Chloe Ballast\n(relay)", 
                "John Windward\n(terminal)"),
  x         = c(-400, 0, 400),     
  y         = c(0, 0, 0),
  color     = c("#F28E2B", "#F28E2B", "#E15759"),
  font.size = 15,
  shape     = "ellipse",
  size      = 38
)

chain_vis_edges <- tibble(
  from      = c(1, 2),
  to        = c(2, 3),
  label     = c("read_file\n11:12:24 UTC", 
                "read_file\n11:21:13 UTC ⚠"),
  arrows    = "to",
  color     = c("grey50", "#E15759"),
  width     = c(2, 3)
)

visNetwork(
  chain_vis_nodes,
  chain_vis_edges,
  main   = list(
    text  = "Figure 6.2: Confirmed Task Delegation Chain — 2046-05-17",
    style = "font-size:14px; font-weight:bold; text-align:center;"
  ),
  width  = "100%",
  height = "280px"
) %>%
  visEdges(
    smooth = list(enabled = FALSE),
    font   = list(size = 16, align = "top")  
  ) %>%
  visPhysics(enabled = FALSE) %>%
  visInteraction(dragNodes = FALSE, zoomView = FALSE, selectable = FALSE)
```

::: callout-note
## Observations

- The chain confirms that **John Windward was the terminal executor**, not the originator, of the task that coincided with the anomalous post at 11:21 UTC.
- **Daniel Gangway** is the earliest identified actor in the chain. Whether he was himself acting on an external instruction cannot be determined from task events alone.
- The nine-minute interval between each delegation step (11:12 → 11:21) is consistent with a deliberate, sequential relay rather than concurrent or accidental co-occurrence.
- Both Chloe Ballast and Daniel Gangway now become subjects of investigation. Their organisational positions and communication patterns around May 17 are examined in Section 6.3.
:::

### 6.3 Communication Network

Identifying the three-person delegation chain in Section 6.2 reframes the investigation. John Windward is no longer the primary subject — he is the terminal executor of an instruction that originated with Daniel Gangway and was relayed through Chloe Ballast. The next question is whether this chain reflects an isolated incident or a pattern of interaction that preceded and surrounded the anomalous post. To answer this, all recorded events involving any of the three agents are extracted from a 48-hour window centred on the incident (May 16–18 2046) and visualised as a communication network.

`visNetwork` is employed here because the resulting network is small enough that all nodes and labels are legible, and the interactive tooltips allow individual edges — including their event type and exact timestamp — to be inspected without cluttering the static plot with text. Edge colour encodes event class to distinguish directive events (`queue_subordinate_task`) from informational ones (file reads, checks), while edge width encodes interaction frequency. The 48-hour window is deliberately narrow: scoping to May 16–18 surfaces only contextually relevant interactions while excluding unrelated activity from the broader dataset.

```{r}
#| code-fold: true
#| label: comms-network
#| code-summary: "Show code"
#| warning: false
#| freeze: false

chain_agents <- c("Agent/person:john_windward",
                  "Agent/person:chloe_ballast",
                  "Agent/person:daniel_gangway")

net_events <- mc2_df %>%
  filter(
    date >= as.Date("2046-05-16"),
    date <= as.Date("2046-05-18")
  ) %>%
  filter(map_lgl(parties, ~any(.x %in% chain_agents)))

net_edges_raw <- net_events %>%
  mutate(
    edge_from   = map_chr(parties, ~.x[[1]]),
    edge_to     = if_else(
      short_name == "queue_subordinate_task" & !is.na(target_agent),
      target_agent,
      map_chr(parties, ~ifelse(length(.x) >= 2, .x[[2]], NA_character_))
    ),
    event_class = if_else(
      short_name == "queue_subordinate_task",
      "Directive", "Informational"
    ),
    tooltip_text = paste0(short_name, "\n", datetime)
  ) %>%
  filter(!is.na(edge_to), edge_from != edge_to)

net_edges_agg <- net_edges_raw %>%
  group_by(edge_from, edge_to, event_class) %>%
  summarise(
    n            = n(),
    edge_tooltip = paste(tooltip_text, collapse = "\n"),
    .groups      = "drop"
  ) %>%
  mutate(
    color = if_else(event_class == "Directive", "#E15759", "#AEC6CF"),
    width = scales::rescale(n, to = c(1, 6))
  )

net_nodes <- tibble(agent = unique(c(net_edges_agg$edge_from,
                                     net_edges_agg$edge_to))) %>%
  mutate(
    id        = row_number(),
    label     = agent %>%
                  str_remove("Agent/person:") %>%
                  str_replace_all("_", " ") %>%
                  str_to_title(),
    color     = case_when(
      agent == "Agent/person:john_windward"                             ~ "#E15759",
      agent %in% c("Agent/person:chloe_ballast",
                   "Agent/person:daniel_gangway")                       ~ "#F28E2B",
      TRUE                                                              ~ "#AEC6CF"
    ),
    font.size = if_else(agent %in% chain_agents, 14, 11),
    size      = if_else(agent %in% chain_agents, 30, 20)
  )

id_map <- net_nodes %>% select(agent, id)

vis_edges <- net_edges_agg %>%
  left_join(id_map, by = c("edge_from" = "agent")) %>% rename(from = id) %>%
  left_join(id_map, by = c("edge_to"   = "agent")) %>% rename(to   = id) %>%
  mutate(
    arrows = "to",
    title  = edge_tooltip
  )

set.seed(99)

graph_obj <- tbl_graph(
  nodes    = net_nodes %>% select(id, label, color, size),
  edges    = vis_edges %>% select(from, to, event_class, color, width),
  directed = TRUE
)

ggraph(graph_obj, layout = "fr") +
geom_edge_link(
    aes(colour = event_class, width = width),
    arrow   = arrow(length = unit(4, "mm"), type = "closed"),
    end_cap = circle(6, "mm"),
    alpha   = 0.75
  ) +
  geom_node_point(
    aes(colour = color),
    size = 7
  ) +
  geom_node_point(
    data = function(x) filter(x, color %in% c("#E15759", "#F28E2B")),
    aes(colour = color),
    size = 12
  ) +
  geom_node_text(
    data       = function(x) filter(x, !color %in% c("#E15759", "#F28E2B")),
    aes(label  = label),
    size       = 2.8,
    colour     = "grey30",
    repel      = TRUE,
    nudge_y    = 0.12,
    bg.colour  = "white",
    bg.r       = 0.12
  ) +
  geom_node_text(
    data       = function(x) filter(x, color %in% c("#E15759", "#F28E2B")),
    aes(label  = label),
    size       = 3.4,
    fontface   = "bold",
    colour     = "grey10",
    repel      = TRUE,
    nudge_y    = 0.18,
    bg.colour  = "white",
    bg.r       = 0.18
  ) +
  scale_edge_colour_manual(
    values = c("Directive" = "#E15759", "Informational" = "#9BBDD4"),
    name   = "Event type"
  ) +
  scale_edge_width(range = c(0.4, 2), guide = "none") +
  scale_colour_identity() +
  labs(
    title   = "Figure 6.3: Communication Network — May 16–18 2046",
    caption = "Red = John Windward  |  Orange = Chloe Ballast & Daniel Gangway  |  Blue = other agents\nRed edges = directive (queue_subordinate_task)  |  Blue edges = informational"
  ) +
  theme_graph(base_size = 11, background = "white") +
  theme(
    plot.title   = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.caption = element_text(size = 8.5, colour = "grey50", hjust = 0.5),
    legend.position = "bottom",
    legend.title    = element_text(size = 10),
    legend.text     = element_text(size = 9)
  )
```

::: callout-note
## Observations

- The **confirmed delegation chain** — Daniel Gangway → Chloe Ballast → John Windward — is visible as the two red directive edges crossing the centre of the network, connecting the three highlighted nodes. No other `queue_subordinate_task` edges appear in the 48-hour window, confirming this chain was a targeted, isolated sequence rather than routine task delegation.

- **Daniel Gangway is the highest-degree node** among human agents, with informational edges extending to seven peripheral agents: Zoey Drydock, Alexander Jetty, Owen Hatch, Lily Anchorline, Lucas Compass, Hannah Starboard, and Gabriel Sonar. This positions him as a coordination hub — his interaction volume is disproportionately large relative to his structural role as the chain's originator.

- **John Windward's edges connect exclusively to system nodes** — `System:Saidit` and `System:File System` — rather than to other human agents. This confirms that John's role in the 48-hour window was purely executor: he interacted with external systems (posting and file operations) but initiated no human-to-human communication of his own.

- **Chloe Ballast's peripheral connections** (Michael Capstan, David Breakwater, Grace Lifeboat, Evelyn Dock) are all informational. These do not appear to be part of the incident chain, but their presence indicates Chloe was operationally active during the window — she was not solely relaying Daniel's instruction.

- Together, the network structure supports the interpretation that **Daniel Gangway initiated and orchestrated the chain**, Chloe Ballast served as an intermediary conduit, and John Windward was the unwitting terminal executor whose agent account was used to publish the anomalous post.
:::

### 6.4 Tracing the Origin of Post Content

Each anomalous post sourced its content through the `content_source` field rather than the normal file pathway — `HiddenOrca.txt` on May 10, `MellowOtter.txt` on May 11, and `SwiftWren.txt` on May 17. To determine where these files originated within TenantThread's system, two searches are performed against the full event log. The first checks whether any `read_file` event was executed by an actor carrying these filenames in their parties record. The second checks whether these filenames appear anywhere in the `file` field across all event types. If either search returns results, the files can be traced to a registered system action. If both return zero rows, the files existed entirely outside the logged audit trail.

```{r}
#| label: content-origin-search
#| code-fold: true
#| code-summary: "Show code"

search1 <- mc2_df %>%
  filter(
    short_name == "read_file",
    map_lgl(parties, ~any(str_detect(.x,
      "HiddenOrca|MellowOtter|SwiftWren")))
  ) %>%
  mutate(parties_str = map_chr(parties,
                               ~paste(.x, collapse = ", "))) %>%
  select(datetime, date, short_name, file,
         content_source, parties_str) %>%
  arrange(datetime)

search2 <- mc2_df %>%
  filter(
    str_detect(file, "HiddenOrca|MellowOtter|SwiftWren")
  ) %>%
  mutate(parties_str = map_chr(parties,
                               ~paste(.x, collapse = ", "))) %>%
  select(datetime, date, short_name, file,
         content_source, parties_str) %>%
  arrange(datetime)

# Search 1 result
if (nrow(search1) == 0) {
  kable(
    data.frame(Result = "0 rows returned — no read_file events reference these filenames in parties."),
    caption = "Search 1: read_file events referencing HiddenOrca / MellowOtter / SwiftWren in parties"
  )
} else {
  kable(search1,
        caption = "Search 1: read_file events referencing filenames in parties")
}

# Search 2 result
if (nrow(search2) == 0) {
  kable(
    data.frame(Result = "0 rows returned — filenames do not appear in any file field."),
    caption = "Search 2: Any event with HiddenOrca / MellowOtter / SwiftWren in file field"
  )
} else {
  kable(search2,
        caption = "Search 2: Any event with filename in file field")
}
```

::: callout-note
## Observations — Content Origin Search

- Both searches return zero rows. The filenames `HiddenOrca.txt`, `MellowOtter.txt`, and `SwiftWren.txt` do not appear anywhere in `mc2_df` outside of the `content_source` field at the moment of posting. They are never registered as file objects, never explicitly read under their own names, and never referenced in any other logged system action across the full event log spanning May to July 2046.

- The content origin is unrecoverable by design. The files were never formally created, stored, or accessed through any auditable pathway within TenantThread's system. Whatever mechanism introduced them into the `content_source` field left no traceable record. This is not a gap in the investigation - it is the forensic conclusion. The operation was deliberately structured to ensure the content source could not be reconstructed after the fact.

- The meaning and probable content of the three files cannot be determined from the available data. The event log records that content was posted, but not what that content said. The immediate deletion of files at +2s and +3s after each post further eliminates any possibility of recovery from within the system logs.
:::

::: callout-warning
## Caveat — Question 2 Partially Unanswerable

The VAST Challenge MC2 Task 2 asks what the posts "mean" and where their content originates. The available data does not contain the text of the posts, and the source files exist nowhere in the event log outside the `content_source` field. The forensic conclusion - that the content was deliberately kept off the audit trail - is itself the answer to the origin question. The meaning of the content remains outside the scope of what `mc2_df` can establish.
:::

## 7. Tracing the Origin of SwiftWren.txt

Sections 5 and 6 established **who** was involved and **how** the task reached John Windward. The outstanding question is: **where did the content of `SwiftWren.txt` come from?** A direct file search yields no results — `SwiftWren.txt` does not appear in the `file` field anywhere across the 185,147 records in the dataset. It surfaces in exactly one row: the anomalous `saidit_post` at `2046-05-17 11:21:15 UTC`, where it is referenced as `content_source`. Because no file object was ever formally registered, the origin must be reconstructed indirectly — by pulling every logged event from the moment the chain was set in motion through to the post and its immediate aftermath, and then examining what destruction activity followed to erase the trail.

### 7.1 Reconstructing the Full Incident Sequence

Section 6.2 confirmed that the task chain was initiated by `daniel_gangway` at `11:12:24 UTC` — nine minutes before the anomalous post. Any investigation window that starts later than this will miss the origin event and present an incomplete causal picture. The query below therefore spans `11:10:00` to `11:22:00 UTC`, covering the full twelve-minute window from safely before Daniel's initiation through to the post and any immediate deletion events.

A sequenced event table is the appropriate display here. The number of events in a twelve-minute window is small enough that individual rows are meaningful, and the tabular format preserves the precise timestamp ordering that is central to the causal argument — something a chart would obscure.

```{r}
#| label: incident-sequence
#| code-fold: true
#| code-summary: "Show code"

incident_sequence <- mc2_df %>%
  filter(
    date     == as.Date("2046-05-17"),
    datetime >= as.POSIXct("2046-05-17 11:10:00", tz = "UTC"),
    datetime <= as.POSIXct("2046-05-17 11:22:00", tz = "UTC")
  ) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = ", "))) %>%
  select(datetime, short_name, file, content_source,
         parties_str, target_agent) %>%
  arrange(datetime)

incident_sequence %>%
  kable(
    col.names = c("Timestamp (UTC)", "Event Type", "File",
                  "Content Source", "Parties", "Target Agent"),
    caption   = "Table 7.1: Event sequence — 11:10 to 11:22 UTC, 17 May 2046"
  )
```

::: callout-note
## Observations — Full Incident Sequence

- The window contains **9 events** across the twelve-minute span. The earliest event is `daniel_gangway`'s `queue_subordinate_task` at `11:12:24` — no events appear before this timestamp within the window, meaning no upstream instruction to Daniel is visible here. Whether Daniel acted on instruction or self-initiated will be examined in Section 7.2.

- The sequence unfolds in four distinct phases. First, `daniel_gangway` delegates a `read_file` task to `chloe_ballast` at `11:12:24`. Nine minutes later, `chloe_ballast` issues a second `queue_subordinate_task` to `john_windward` at `11:21:13` - confirmed in Section 6.2 as carrying the same `read_file` task type, establishing Chloe as a relay node rather than an originating actor. One second later, a **`saidit_post_check`** event fires at `11:21:14` — a pre-post validation step — which failed to detect the anomalous `content_source`. One second after that, at `11:21:15`, the anomalous post is published with `content_source = SwiftWren.txt`.

- **`SwiftWren.txt` appears exclusively in the `content_source` column** — the `file` field is `NA` across every row in the table, confirming that the file was never registered as a formal file object in the system. It was consumed as read content and injected directly into the post body.

- Two **`delete_file`** events follow at `11:21:16` and `11:21:17` — one and two seconds after the post respectively — both attributed to `john_windward` and `system:file_system`. The `file` field is `NA` for both deletions, meaning the filenames cannot be recovered from this table alone. Section 7.3 targets these deletions directly to establish what was destroyed.

- Two unrelated `read_file` events by `samuel_gull` (`11:17:45`) and `michael_capstan` (`11:19:55`) appear in the window. Neither is linked to the task chain by `parties` or `target_agent` and are treated as coincidental background activity. Similarly, `grace_lifeboat`'s `read_file` at `11:21:50` shows no connection to the chain.
:::

### 7.2 Tracing the Upstream of Daniel Gangway

Table 7.1 showed no events before `11:12:24` within the incident window, but that window was sized to capture the incident sequence - not to establish what preceded Daniel. To determine whether Daniel acted on instruction or self-initiated, all events involving `daniel_gangway` on May 17 2046 prior to `11:12:24 UTC` are examined. The results are ordered in reverse chronological order to surface events closest to the initiation first.

```{r}
#| label: daniel-upstream
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(
    date     == as.Date("2046-05-17"),
    datetime <  as.POSIXct("2046-05-17 11:12:24", tz = "UTC"),
    map_lgl(parties, ~any(str_detect(.x, "daniel_gangway")))
  ) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = ", "))) %>%
  select(datetime, short_name, file, content_source,
         parties_str, target_agent) %>%
  arrange(desc(datetime)) %>%
  head(20) %>%
  kable(
    col.names = c("Timestamp (UTC)", "Event Type", "File",
                  "Content Source", "Parties", "Target Agent"),
    caption   = "Table 7.2a: Events involving daniel_gangway before 11:12:24 UTC"
  )
```

The most proximate event to Daniel's initiation is a `queue_subordinate_task` from `lily_anchorline` at `10:11:31 UTC` - targeting Daniel directly. To confirm whether this instruction is the origin of the `read_file` chain that ultomately produced the anomalous post, the full event record at that timestamp is inspected, including the `task` field.

```{r}
#| label: lily-task-confirm
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(
    date     == as.Date("2046-05-17"),
    datetime == as.POSIXct("2046-05-17 10:11:31", tz = "UTC"),
    map_lgl(parties, ~any(str_detect(.x, "lily_anchorline")))
  ) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = ", "))) %>%
  select(datetime, short_name, task, file, content_source,
         parties_str, target_agent) %>%
  kable(
    col.names = c("Timestamp (UTC)", "Event Type", "Task", "File",
                  "Content Source", "Parties", "Target Agent"),
    caption   = "Table 7.2b: Full event record — lily_anchorline instruction at 10:11:31 UTC"
  )
```

::: callout-note
## Observations — Upstream of Daniel Gangway

- Table 7.2a returns **17 events** involving `daniel_gangway` on May 17 2046 prior to `11:12:24 UTC`. The majority are routine `check_email` events between `07:55` and `08:02` — background activity unrelated to the incident.

- **The critical finding is in Table 7.2a row 2**: at `10:11:31 UTC` — exactly one hour before Daniel initiated the task chain — `lily_anchorline` issued a `queue_subordinate_task` directly to `daniel_gangway`, confirmed by `target_agent = Agent/person:daniel_gangway`. Daniel is not the originating actor. He was acting on instruction.

- **Table 7.2b confirms the relevance of this instruction**: The `task` field records `read_file` - the same task type that Daniel delegated to `chloe_ballast` at `11:12:24`, and that Chloe in turn delegated to `john_windward` at `11:21:13`. The chain is now traceable to four actors: **`lily_anchorline` -\> `daniel_gangway` -\> `chloe_ballast` -\> `john_windward`**.

- One second after receiving lily's instruction, at `10:11:32`, Daniel ran a **`saidit_post_check`** — the same pre-post validation event observed at `11:21:14` immediately before the anomalous post. A post check at this point suggests Daniel was assessing whether a post could be made, but no post followed.

- **The gap between lily's instruction (`10:11:31 UTC`) and Daniel's first delegation (`11:12:24 UTC`) spans one hour and one minute.** A targeted query across this interval returns **zero events** involving `daniel_gangway` - he is completely absent from the system logs during this period. No intermediate steps, file reads, or communications are recorded. Whether Daniel was processing the instruction offline or awaiting a signal cannot be determined from teh available logs. This interval is noted as an unsolved gap in the evidence chain.

- With `lily_anchorline` confiremed as the actor directly above Daniel, the investigation traces one step further: who instructed Lily?\
:::

The chain above Daniel extends at least one level higher. The next step applies the same upstream tracing method to `lily_anchorline` — examining all events involving Lily on May 17 prior to her `10:11:31` delegation to Daniel — to determine whether she too was acting on instruction.

```{r}
#| label: lily-upstream
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(
    date     == as.Date("2046-05-17"),
    datetime <  as.POSIXct("2046-05-17 10:11:31", tz = "UTC"),
    map_lgl(parties, ~any(str_detect(.x, "lily_anchorline")))
  ) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = ", "))) %>%
  select(datetime, short_name, task, file, content_source,
         parties_str, target_agent) %>%
  arrange(desc(datetime)) %>%
  head(10) %>%
  kable(
    col.names = c("Timestamp (UTC)", "Event Type", "Task", "File",
                  "Content Source", "Parties", "Target Agent"),
    caption   = "Table 7.2c: Events involving lily_anchorline before 10:11:31 UTC"
  )
```

Table 7.2c returns 2 rows. The `check_email` at `10:09:47` is routine background activity. The critical row is the `queue_subordinate_task` at `08:16:07 UTC` — `zoey_drydock` and `lily_anchorline` appear together in parties, and `target_agent = Agent/person:lily_anchorline` confirms that Zoey was the issuer and Lily the receiver, carrying a `read_file` instruction. Lily was herself acting on instruction, passed down from `zoey_drydock` more than two hours before she instructed Daniel. To surface the full relay structure in one view, all `queue_subordinate_task` events on May 17 preceding the anomalous post are assembled chronologically. This makes any looping or circular routing immediately visible without running a separate upstream query at each step.

```{r}
#| label: full-chain-reconstruction
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 14
#| fig-height: 6
#| warning: false

relay_nodes <- tibble(
  name = c("james_stern", "gabriel_sonar", "daniel_gangway",
           "zoey_drydock", "mia_fender", "victoria_rigging",
           "lily_anchorline", "chloe_ballast", "john_windward"),
  label    = str_replace_all(name, "_", " ") %>% str_to_title(),
  is_reused = name %in% c("daniel_gangway", "zoey_drydock")
)

relay_edges <- tibble(
  from = c("james_stern", "gabriel_sonar", "daniel_gangway",
           "zoey_drydock", "mia_fender", "victoria_rigging",
           "zoey_drydock", "lily_anchorline", "daniel_gangway",
           "chloe_ballast"),
  to   = c("gabriel_sonar", "daniel_gangway", "zoey_drydock",
           "mia_fender", "victoria_rigging", "zoey_drydock",
           "lily_anchorline", "daniel_gangway", "chloe_ballast",
           "john_windward"),
  time        = c("00:51", "02:26", "04:12", "05:34", "05:52",
                  "06:44", "08:16", "10:11", "11:12", "11:21"),
  is_terminal = c(rep(FALSE, 9), TRUE)
)

relay_graph <- tbl_graph(
  nodes    = relay_nodes,
  edges    = relay_edges,
  directed = TRUE
)

ggraph(relay_graph, layout = "linear") +
  geom_edge_arc(
    aes(colour = is_terminal,
        label  = time),
    arrow        = arrow(length = unit(3, "mm"), type = "closed"),
    end_cap      = circle(5, "mm"),
    strength     = 0.4,
    alpha        = 0.8,
    label_size   = 3,
    label_colour = "grey40",
    label_dodge  = unit(4, "mm"),
    angle_calc   = "along",
    force_flip   = TRUE
  ) +
  geom_node_point(
    aes(colour = is_reused),
    size = 6
  ) +
  geom_node_text(
    aes(label = label),
    size    = 3,
    nudge_y = -0.18
  ) +
  scale_edge_colour_manual(
    values = c("FALSE" = "#4E79A7", "TRUE" = "#E15759"),
    guide  = "none"
  ) +
  scale_colour_manual(
    values = c("FALSE" = "#4E79A7", "TRUE" = "#F28E2B"),
    labels = c("Single appearance", "Reused node"),
    name   = "Node type"
  ) +
  labs(
    title    = "Figure 7.2d: Relay Chain — May 17 2046",
    subtitle = "Orange = reused nodes (Daniel Gangway, Zoey Drydock)  |  Red arc = terminal delegation  |  Labels show delegation time (UTC)"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 9, colour = "grey50"),
    legend.position = "bottom",
    legend.text     = element_text(size = 9),
    legend.title    = element_text(size = 9)
  )
```

::: callout-note
## Observations — Full Relay Chain Reconstruction

- **Figure 7.2d shows 10 delegation events**, tracing the full relay chain from its earliest visible actor to the terminal executor. Read chronologically, the sequence is:

- 00:51:43 `james_stern` → `gabriel_sonar` (read_file)

- 02:26:11 `gabriel_sonar` → `daniel_gangway` (read_file)

- 04:12:32 `daniel_gangway` → `zoey_drydock` (read_file)

- 05:34:32 `zoey_drydock` → `mia_fender` (read_file)

- 05:52:51 `mia_fender` → `victoria_rigging` (read_file)

- 06:44:28 `victoria_rigging` → `zoey_drydock` (read_file)

- 08:16:07 `zoey_drydock` → `lily_anchorline` (read_file)

- 10:11:31 `lily_anchorline` → `daniel_gangway` (read_file)

- 11:12:24 `daniel_gangway` → `chloe_ballast` (read_file)

- 11:21:13 `chloe_ballast` → `john_windward` (read_file) → **anomalous post**

- Two actors appear at more than one position in the chain. `daniel_gangway` receives a task from `gabriel_sonar` at `02:26:11`, issues one to `zoey_drydock` at `04:12:32`, receives again from `lily_anchorline` at `10:11:31`, and issues again to `chloe_ballast` at `11:12:24`. `zoey_drydock` similarly receives from `daniel_gangway` at `04:12:32`, issues to `mia_fender` at `05:34:32`, receives again from `victoria_rigging` at `06:44:28`, and issues again to `lily_anchorline` at `08:16:07`. This is not a simple linear relay — the same agents are reused at non-adjacent positions, meaning a back-trace that stops when it first encounters a known actor will terminate prematurely and misidentify that actor as the originator.

- **`james_stern` is the earliest actor visible in the log at `00:51:43`**. Whether he was himself acting on instruction cannot be determined from `mc2_df` — no prior event involving `james_stern` on May 17 in connection with this chain is visible. This is acknowledged as the chain limit of the available evidence, not a confirmed point of origin.

- **The entire operation spans ten hours and twenty-nine minutes**, from `james_stern`'s first delegation at `00:51:43` to `chloe_ballast`'s final handoff to `john_windward` at `11:21:13` — a sustained, coordinated sequence executed across the full morning of May 17 2046.
:::

### 7.3 Scoping the Cover-Up - Deletion of Events

Two `delete_file` events were recorded at `11:21:16` and `11:21:17` - one and two seconds after the anomalous post, and both attributed to `john_windward` and `system:file_system`. In Table 7.1, the `file` field appeared as `NA` for both rows. A dedicated query targetting all `delete_file` events attributed to `john_windward` on May 17 is run to determine whether a broader query context surfaces filenames that the incident-sequence filter did not expose.

```{r}
#| label: delete-events
#| code-fold: true
#| code-summary: "Show code"

mc2_df %>%
  filter(
    short_name == "delete_file",
    date       == as.Date("2046-05-17"),
    map_lgl(parties, ~any(str_detect(.x, "john_windward")))
  ) %>%
  mutate(parties_str = map_chr(parties, ~paste(.x, collapse = ", "))) %>%
  select(datetime, short_name, file, parties_str) %>%
  arrange(datetime) %>%
  kable(
    col.names = c("Timestamp (UTC)", "Event Type", "File", "Parties"),
    caption   = "Table 7.3: delete_file events attributed to john_windward — May 17 2046"
  )
```

::: callout-note
## Observations — Deletion Events

- Table 7.3 confirms the two `delete_file` events at `11:21:16` and `11:21:17` identified in Table 7.1. The `file` field is `NA` in both rows even under this dedicated query — the filenames of the destroyed objects are not recoverable from the event log. The system recorded that deletions occurred, but not what was deleted.

- **The timing is unambiguous**. Both deletions execute within two seconds of the anomalous post at 11:21:15, consistent with a scripted sequence in which publishing and evidence destruction were coordinated as a single operation rather than independent actions.

- **Two separate deletion events in consecutive seconds indicate two distinct file objects were destroyed**. Their contents cannot be established from the available data.
:::

::: callout-warning
## Caveat — Irrecoverable Filenames

The `file` field is `NA` for both deletion events under all query conditions tested. No alternative field in `mc2_df` contains the filenames of the deleted objects. The identity of the destroyed files cannot be established from the available data.
:::

### 7.4 Synthesis - The Full Incident Timeline

The preceding sub-sections established the delegation chain, the relay structure, the failed validation, and the post-publication cleanup as separate findings. Figure 7.4 will consoolidate all these into a single visual - plotting every relevant event from `00:51:43` to `11:21:17 UTC` across all actors involved, coloured by event type. This allows the full arc of the incident to be read in one view - the relay building across the morning, the behavioural signature repeating at each node, and the abrupt post-and-delete sequence at the terminal executor.

```{r}
#| label: incident-timeline
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 13
#| fig-height: 7

chain_actors <- c("james_stern", "gabriel_sonar", "daniel_gangway",
                  "zoey_drydock", "mia_fender", "victoria_rigging",
                  "lily_anchorline", "chloe_ballast", "john_windward")

timeline_events <- mc2_df %>%
  filter(
    date == as.Date("2046-05-17"),
    datetime <= as.POSIXct("2046-05-17 11:21:17", tz = "UTC"),
    short_name %in% c("queue_subordinate_task", "saidit_post_check",
                      "saidit_post", "delete_file"),
    map_lgl(parties, ~any(
      str_remove(.x, "Agent/person:") %in% chain_actors |
      str_detect(.x, "john_windward")
    ))
  ) %>%
  mutate(
    actor = map_chr(parties, ~{
      hit <- str_remove(.x, "Agent/person:")[
        str_remove(.x, "Agent/person:") %in% chain_actors]
      if (length(hit) > 0) hit[[1]] else NA_character_
    }),
    event_type = factor(short_name,
                        levels = c("queue_subordinate_task",
                                   "saidit_post_check",
                                   "saidit_post",
                                   "delete_file"),
                        labels = c("Delegation",
                                   "Post Check",
                                   "Anomalous Post",
                                   "Deletion")),
    tooltip = case_when(
      short_name == "queue_subordinate_task" ~ paste0(
        "Delegation\n",
        str_replace_all(map_chr(parties, ~str_remove(.x[[1]], "Agent/person:")), "_", " ") %>% str_to_title(),
        " → ",
        str_replace_all(str_remove(target_agent, "Agent/person:"), "_", " ") %>% str_to_title(),
        "\nTime: ", format(datetime, "%H:%M:%S UTC")
      ),
  TRUE ~ paste0(
    "Actor: ", str_replace_all(actor, "_", " ") %>% str_to_title(), "\n",
    "Event: ", event_type, "\n",
    "Time: ", format(datetime, "%H:%M:%S UTC")
  )
),
  ) %>%
  filter(!is.na(actor))

p <- ggplot(timeline_events,
       aes(x      = datetime,
           y      = factor(actor,
                           levels = rev(chain_actors),
                           labels = rev(str_replace_all(chain_actors, "_", " ") %>%
                                          str_to_title())),
           colour = event_type,
           shape  = event_type,
           text   = tooltip)) +
  geom_hline(yintercept = seq_along(chain_actors),
             colour = "grey92", linewidth = 0.4) +
  geom_vline(xintercept = as.POSIXct("2046-05-17 11:21:15", tz = "UTC"),
             linetype = "dashed", colour = "#E15759", linewidth = 0.6) +
  geom_point(data = ~filter(.x, event_type != "Deletion"),
             size = 4, alpha = 0.9) +
  geom_point(data = ~filter(.x, event_type == "Deletion"),
             size = 4, alpha = 0.9,
             position = position_jitter(height = 0.15, width = 0)) +
  scale_colour_manual(
    values = c("Delegation"     = "#4E79A7",
               "Post Check"     = "#F28E2B",
               "Anomalous Post" = "#E15759",
               "Deletion"       = "#555555"),
    name = "Event Type"
  ) +
  scale_shape_manual(
    values = c("Delegation"     = 16,
               "Post Check"     = 17,
               "Anomalous Post" = 8,
               "Deletion"       = 4),
    name = "Event Type"
  ) +
  scale_x_datetime(date_labels = "%H:%M",
                   date_breaks = "1 hour") +
  labs(
    title = "Figure 7.4: Full Incident Timeline — May 17 2046",
    x     = "Time (UTC)",
    y     = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    panel.grid    = element_blank(),
    axis.text.y   = element_text(size = 10),
    axis.text.x   = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggplotly(p, tooltip = "text") %>%
  layout(
    title  = list(text = "Figure 7.4: Full Incident Timeline — May 17 2046",
                  font = list(size = 14)),
    margin = list(r = 80, b = 140),
    xaxis  = list(title = list(text  = "Time (UTC)",
                               standoff = 30)),
    legend = list(orientation = "h",
                  y = -0.25,
                  x = 0.5,
                  xanchor = "center")
  )
```

::: callout-note
## Observations — Full Incident Timeline

- **Figure 7.4 consolidates all 15 events** across the nine confirmed chain actors into a single view, spanning `00:51:43` to `11:21:17 UTC`. The progression of blue delegation dots from top-left to bottom-right traces the relay passing sequentially down the chain across the full morning.

- **The post-check behavioural signature is visible on two relay nodes**. The orange triangles on `zoey_drydock` (\~`06:44`) and `daniel_gangway` (\~`10:11`) mark `saidit_post_check` events — fired immediately after each received a delegation — before passing the task further down. This repeating pattern across non-adjacent actors is consistent with a coordinated scripted protocol rather than independent behaviour.

- **The dashed red line marks the moment of convergence at `11:21:15 UTC`**. All activity in the preceding ten hours leads to this point. At `john_windward`'s row, the anomalous post (asterisk) and two deletion events (X) are clustered within two seconds of each other — the post-and-destroy sequence executing as a single coordinated operation. Hovering over each symbol confirms the precise timestamp and event type.

- **`james_stern` and `gabriel_sonar` each have only one event visible**, both delegations, with no post-check signature. Whether this reflects a different role in the operation or simply a gap in the available log cannot be determined from the current evidence.
:::

## 8. Historical Anomalies - Recurrence Analysis

Section 5 established that three anomalous `saidit_post` events exist in the dataset — `HiddenOrca.txt` (May 10), `MellowOtter.txt` (May 11), and `SwiftWren.txt` (May 17). Section 7 reconstructed the full chain behind the May 17 incident in detail. Before proposing an intervention, a critical question must be answered: were these three incidents independent occurrences, or manifestations of the same coordinated operation? This section examines two dimensions of recurrence — actor participation across all three dates, and the behavioural signature at the terminal executor — to determine whether a common underlying operation connects them.

### 8.1 Actor Participation Across Incidents

The first dimension of recurrence examined is actor participation. If the same agents appear across all three incidents, it suggests a shared network rather than three independent chains. All `queue_subordinate_task` (read_file) delegation events on each incident date are extracted up to the moment of the anomalous post, and the unique actors across issuer and receiver roles are identified. Figure 8.1 presents the result as a heatmap — actors on the y-axis ordered by number of incidents they appear in, incident dates on the x-axis — making the overlap immediately visible.

```{r}
#| label: recurrence-heatmap
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 11
#| fig-height: 6

get_chain_actors <- function(target_date, cutoff_time) {
  mc2_df %>%
    filter(
      date       == as.Date(target_date),
      short_name == "queue_subordinate_task",
      task       == "read_file",
      datetime   <= as.POSIXct(cutoff_time, tz = "UTC")
    ) %>%
    mutate(
      issuer   = map_chr(parties, ~str_remove(.x[[1]], "Agent/person:")),
      receiver = str_remove(target_agent, "Agent/person:")
    ) %>%
    select(issuer, receiver) %>%
    pivot_longer(everything(), values_to = "actor") %>%
    distinct(actor) %>%
    filter(!is.na(actor)) %>%
    mutate(date = target_date)
}

chain_may10 <- get_chain_actors("2046-05-10", "2046-05-10 12:45:40")
chain_may11 <- get_chain_actors("2046-05-11", "2046-05-11 13:03:30")
chain_may17 <- get_chain_actors("2046-05-17", "2046-05-17 11:21:13")

all_chains <- bind_rows(chain_may10, chain_may11, chain_may17) %>%
  mutate(
    present = 1,
    date    = factor(date,
                     levels = c("2046-05-10", "2046-05-11", "2046-05-17"),
                     labels = c("May 10\n(HiddenOrca.txt)",
                                "May 11\n(MellowOtter.txt)",
                                "May 17\n(SwiftWren.txt)"))
  )

actor_freq <- all_chains %>%
  group_by(actor) %>%
  summarise(n_incidents = n_distinct(date), .groups = "drop") %>%
  arrange(desc(n_incidents), actor)

heatmap_data <- all_chains %>%
  complete(actor, date, fill = list(present = 0)) %>%
  left_join(actor_freq, by = "actor") %>%
  mutate(
    actor = str_replace_all(actor, "_", " ") %>% str_to_title(),
    fill_label = if_else(present == 1, "Present", "Absent"),
    is_terminal = str_detect(actor, "John Windward")
  )

ggplot(heatmap_data,
       aes(x    = date,
           y    = reorder(actor, n_incidents),
           fill = fill_label)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(
    data = filter(heatmap_data, is_terminal),
    aes(label = "★"),
    colour = "white", size = 5
  ) +
  scale_fill_manual(
    values = c("Present" = "#4E79A7", "Absent" = "#E8E8E8"),
    name   = NULL
  ) +
  labs(
    title    = "Figure 8.1: Actor Participation Across Three Anomalous Incidents",
    subtitle = "★ marks John Windward — terminal executor in all three incidents | Actors ordered by number of incidents",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 9, colour = "grey50"),
    panel.grid      = element_blank(),
    axis.text.y     = element_text(size = 9),
    axis.text.x     = element_text(size = 10, face = "bold"),
    legend.position = "bottom"
  )
```

::: callout-note
## Observations — Actor Participation Across Three Anomalous Incidents

- **Figure 8.1 confirms that the three anomalous incidents are not isolated events** — they share a persistent core of actors across all three dates. Six actors appear in every incident: `victoria_rigging`, `mia_fender`, `lily_anchorline`, `john_windward`, `daniel_gangway`, and `chloe_ballast.` Their presence across May 10, May 11, and May 17 is inconsistent with coincidence and consistent with a coordinated, repeated operation using the same underlying network.

- **`john_windward` is the terminal executor in all three incidents**, confirmed by the starred row appearing fully blue across all three columns. On May 10 he received the final delegation from `chloe_ballast` at `12:45:40`; on May 11 from `liam_anchor` at `13:03:30`; and on May 17 from `chloe_ballast` at `11:21:13`. The terminal role is consistently assigned to the same agent regardless of which routing path was used to reach him.

- **`zoey_drydock` is absent from May 10 but present in May 11 and May 17**, suggesting her role in the operation expanded across the three incidents. Her entry into the chain coincides with the transition from the first to the second incident, and she remains a participant through to May 17.

- **Incident-specific actors are visible in the lower rows of Figure 8.1**. `levi_signal`, `michael_capstan`, and `owen_hatch` appear only in May 10; `henry_sail`, `david_breakwater`, and `olivia_keel` only in May 11; `james_stern` only in May 17. These actors represent the variable outer layers of each incident's relay — the routing path changes across incidents, but the core network and terminal executor remain constant.

- The combination of a stable core, a fixed terminal executor, and variable outer routing across three separate dates is the defining **structural signature** of this operation. It indicates that the operation is resilient by design — the outer relay can be changed to vary the apparent chain while the core actors and terminal executor remain unchanged and undetected.
:::

### 8.2 Behaviour Signature at the Terminal Node

Actor overlap alone does not prove the incidents share the same operation — the same agents could participate in unrelated chains. The stronger test is whether the same behavioural sequence executes at the terminal node each time. Section 7 established that on May 17, `john_windward` executed a precise four-event sequence — `saidit_post_check` → `saidit_post` → `delete_file` → `delete_file` — at one-second intervals, with post content delivered exclusively through the `content_source` field rather than the normal file pathway. Figure 8.2 examines whether this identical sequence repeats on May 10 and May 11, normalising each incident to the moment the post check fires to make the timing directly comparable.

```{r}
#| label: terminal-sequence-plot
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 10
#| fig-height: 5

sequence_data <- mc2_df %>%
  filter(
    date %in% as.Date(c("2046-05-10", "2046-05-11", "2046-05-17")),
    short_name %in% c("saidit_post_check", "saidit_post", "delete_file"),
    map_lgl(parties, ~any(str_detect(.x, "john_windward")))
  ) %>%
  mutate(
    incident = case_when(
      date == as.Date("2046-05-10") ~ "May 10\n(HiddenOrca.txt)",
      date == as.Date("2046-05-11") ~ "May 11\n(MellowOtter.txt)",
      date == as.Date("2046-05-17") ~ "May 17\n(SwiftWren.txt)"
    ),
    event_type = factor(short_name,
                        levels = c("saidit_post_check",
                                   "saidit_post",
                                   "delete_file"),
                        labels = c("Post Check",
                                   "Anomalous Post",
                                   "Deletion"))
  ) %>%
  mutate(
    seconds_offset = as.numeric(difftime(
      datetime, min(datetime), units = "secs"
    )),
    .by = date
  ) %>%
  mutate(
    point_label = if_else(
      seconds_offset == 0,
      format(datetime, "%H:%M:%S"),
      paste0("+", seconds_offset, "s")
    )
  )

post_labels <- sequence_data %>%
  filter(event_type == "Anomalous Post") %>%
  mutate(content_label = case_when(
    date == as.Date("2046-05-10") ~ "content_source:\nHiddenOrca.txt",
    date == as.Date("2046-05-11") ~ "content_source:\nMellowOtter.txt",
    date == as.Date("2046-05-17") ~ "content_source:\nSwiftWren.txt"
  ))

ggplot(sequence_data,
       aes(x      = seconds_offset,
           y      = factor(incident,
                           levels = c("May 17\n(SwiftWren.txt)",
                                      "May 11\n(MellowOtter.txt)",
                                      "May 10\n(HiddenOrca.txt)")),
           colour = event_type,
           shape  = event_type)) +
  geom_hline(yintercept = 1:3, colour = "grey92", linewidth = 0.4) +
  geom_point(size = 5, alpha = 0.9) +
  geom_text(aes(label = point_label),
            vjust = -1, size = 3.2, colour = "grey40") +
  # content_source annotation below the anomalous post point
  geom_text(data = post_labels,
            aes(label = content_label),
            vjust = 2.2, size = 2.8,
            colour = "#E15759",
            inherit.aes = TRUE) +
  scale_colour_manual(
    values = c("Post Check"     = "#F28E2B",
               "Anomalous Post" = "#E15759",
               "Deletion"       = "#555555"),
    name = "Event Type"
  ) +
  scale_shape_manual(
    values = c("Post Check"     = 17,
               "Anomalous Post" = 8,
               "Deletion"       = 4),
    name = "Event Type"
  ) +
  scale_x_continuous(
    breaks = 0:3,
    labels = c("Post Check", "+1s", "+2s", "+3s"),
    limits = c(-0.3, 3.8)
  ) +
  labs(
    title    = "Figure 8.2: Terminal Four-Event Sequence — All Three Incidents",
    subtitle = "Actual post check time shown at +0s | content_source field confirmed on each anomalous post",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 9, colour = "grey50"),
    panel.grid      = element_blank(),
    axis.text.y     = element_text(size = 10, face = "bold"),
    axis.text.x     = element_text(size = 10),
    legend.position = "bottom"
  )
```

::: callout-note
## Observations — Behavioural Signature at the Terminal Node

- **Figure 8.2 confirms that the identical four-event sequence** — `saidit_post_check` → `saidit_post` → `delete_file` → `delete_file` — executes at one-second intervals across all three incidents, attributed to `john_windward` each time. The post check fires at `12:45:41` on May 10, `00:56:03` on May 11, and 11:21:14 on May 17 — three different times of day, but the same scripted sequence in each case.

- \*\*The `content_source` field is the exclusive delivery mechanism in all three incidents. The `file` field is `NA` in every `saidit_post` event — `HiddenOrca.txt`, `MellowOtter.txt`, and `SwiftWren.txt` are each delivered through `content_source` only, bypassing the normal file registration pathway and evading standard file-based audit controls each time.

- \*\*The precision of the one-second interval timing across three separate dates rules out manual action. A human operator could not reproduce this exact sequence and timing consistently across incidents separated by days. The terminal sequence is scripted — the same protocol executing at the same node, on instruction, each time it is triggered.

- Taken together, Figures 8.1 and 8.2 establish that the three anomalous incidents are not independent events. They share a stable core actor network, a fixed terminal executor, an identical delivery mechanism, and a scripted terminal sequence. This is a single coordinated operation that has executed at least three times.
:::

::: callout-warning
## Unresolved Anomaly — Sustained Virus Activity on `zoey_drydock`

- During the course of data preparation for Section 8, a separate anomaly was observed in the event log. Starting May 10 at `20:10:04` — immediately following the conclusion of the first anomalous posting incident — `zoey_drydock` is the subject of hundreds of `queue_subordinate_task` events with `task = "virus"`, continuing at five-second intervals through May 11. Over 2,800 such events are recorded across the two

- This activity was not investigated as part of this study, which is scoped to the three anomalous `saidit_post` incidents and their associated delegation chains. However, given that `zoey_drydock` is a confirmed participant in the May 11 and May 17 posting chains, the overlap between her posting chain involvement and this sustained virus activity is noted as a significant unresolved anomaly warranting separate investigation.
:::

## 9. Intervention Proposal

The investigation across Sections 7 & 8 has established that three anomalous `saidit_post` events were executed by the same terminal actor, using the same delivery mechanism, through the same validation gate, on three separate dates. In every case, the existing `saidit_post_check` fired one second before the post was posted, yet failed to prevent it. The checks are not absent, just not enough. The gap is specific and identifiable - the existing validation does not evaluate the `content_source` field, allowing post content to bypass audit controls. This section proposes a single, well-defined intervention at the point wheere the existing gate already sits, requiring no new infrastructure but only additional condition on the check that is already in place.

### 9.1 The Vulnerability

The existing `saidit_post_check` event is a validation gate that fires immediately before every `saidit_post` which is confirmed in all three incidents at `+0s`. Its presence in the log proves the gate is not absent. However, across all three anomalous posts, the check fired and passed without blocking the publication. The reason is structural as the existiing check evalueates whether a post can be made, but does not inspect **how** the post content arrived. Sepcificially, it does not evaluate the `content_source` field.

In a normal posting event, content is delivered through the registered file pathway - the `file` field carries the filename and the `content_source` field is `NA`. In all three anomalous posts, this is reversed, where `file` is `NA` and `content_source` carries the filename. This inversion is the signature of the exploit. Content delivered through `content_source` bypasses file registration, bypass file-based audit controls, and is invisible to any check that only evaulates the `file` field.

**The vulnerability is therefore not a missing gate but a missing condition within the gate that already exists.**

### 9.2 The Proposed Intervention

The intervention requires adding a single condition to the existing `saidit_post_check` validation page: **If `content_source` is not null at the point of `saidit_post_check` -\> block the post and raise an alert**

This rule targets the exact inversion identified in Section 9.1. Under normal posting behaviour, `content_source` is always `NA` - a non-null value in this field is not a legitimate system state. It is unambiguously anomalous. The condition therefore carries no risk of false positives against normal posting operations.

The intervention sits at the existing gate with no new infrastructure, no new event type, and no changes to the delegation or file syustem architecture are required. The `saidit_post_check` already executes at the right moment (one second after every post). Adding `content_source` null-check to its evaluation logic is sufficient to intercept the operation at the point it has consistently exploited across all three incidents.

Had this condition been active on May 10, May 11, and May 17, all three anomalous posts would have been blocked at `+0s`, before the post fired, before the deletions executed, and before the content was published. The one-second window between the check and the post is preceisely where the intervention takes effect.

The proposed rule does not resolve the upstream delegation chain, identify the true originator, or prevent the relay from assembling. It closes the specific gap that allowed the terminal executor to publish - *which is the single intervention point Objective 3b requires*.

### 9.3 Visual - Intervention at the Validation Gate

Figure 9.1 presents the posting sequence as a before and after process flow. The left panel shows the current state, where the existing `saidit_post_check` passes without evaluating `content_source`, allowing the anomalous post to proceed. The right panel shows the proposed state, where a single additional condition blocks any post where `content_source` is no null, raising an alert before publication occurs.

```{r}
#| label: intervention-flow-ggplot
#| code-fold: true
#| code-summary: "Show code"
#| fig-width: 14
#| fig-height: 9

diamond_df <- function(cx, cy, w, h, id) {
  tibble(
    x     = c(cx, cx + w/2, cx, cx - w/2),
    y     = c(cy + h/2, cy, cy - h/2, cy),
    group = id
  )
}

left <- ggplot() +
  annotate("rect", xmin=0.1, xmax=5.9, ymin=0.3, ymax=10.8,
           fill="#FFF5F5", color="#E15759", linewidth=1) +
  annotate("text", x=3, y=10.4,
           label="CURRENT STATE \u2014 Vulnerable",
           color="#E15759", fontface="bold", size=4.5) +
  annotate("rect", xmin=1.3, xmax=4.7, ymin=8.8, ymax=9.8,
           fill="#4E79A7", color="white") +
  annotate("text", x=3, y=9.3,
           label="queue_subordinate_task\n(read_file)",
           color="white", size=3.2) +
  annotate("segment", x=3, xend=3, y=8.8, yend=7.7,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  geom_polygon(data=diamond_df(3, 6.5, 3.6, 2.2, "d1"),
               aes(x=x, y=y, group=group),
               fill="#F28E2B", color="white") +
  annotate("text", x=3, y=6.5,
           label="saidit_post_check\nEvaluates: can post be made?\nDoes NOT check content_source",
           color="white", size=2.8) +
  annotate("text", x=2.2, y=5.3,
           label="PASS", color="#E15759", fontface="bold", size=3.5) +
  annotate("segment", x=3, xend=3, y=5.4, yend=4.5,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  annotate("rect", xmin=0.8, xmax=5.2, ymin=3.4, ymax=4.5,
           fill="#E15759", color="white") +
  annotate("text", x=3, y=3.95,
           label="saidit_post\ncontent_source = HiddenOrca.txt  |  file = NA",
           color="white", size=2.8) +
  annotate("text", x=2.2, y=3.0,
           label="+1s / +2s", color="grey50", size=3) +
  annotate("segment", x=3, xend=3, y=3.4, yend=2.5,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  annotate("rect", xmin=1.3, xmax=4.7, ymin=1.4, ymax=2.5,
           fill="#555555", color="white") +
  annotate("text", x=3, y=1.95,
           label="delete_file \u00d7 2\nEvidence destroyed",
           color="white", size=3) +
  coord_cartesian(xlim=c(0, 6), ylim=c(0.5, 11)) +
  theme_void()

# ── RIGHT PANEL — Proposed State ─────────────────────────────────────────────
right <- ggplot() +
  annotate("rect", xmin=0.1, xmax=7.9, ymin=0.3, ymax=10.8,
           fill="#F5FFF5", color="#59A14F", linewidth=1) +
  annotate("text", x=4, y=10.4,
           label="PROPOSED STATE \u2014 Intervention",
           color="#59A14F", fontface="bold", size=4.5) +
  annotate("rect", xmin=2, xmax=6, ymin=8.8, ymax=9.8,
           fill="#4E79A7", color="white") +
  annotate("text", x=4, y=9.3,
           label="queue_subordinate_task\n(read_file)",
           color="white", size=3.2) +
  annotate("segment", x=4, xend=4, y=8.8, yend=7.7,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  geom_polygon(data=diamond_df(4, 6.5, 3.6, 2.2, "d2"),
               aes(x=x, y=y, group=group),
               fill="#F28E2B", color="white") +
  annotate("text", x=4, y=6.5,
           label="saidit_post_check\nEvaluates: can post be made?\nEvaluates: is content_source NULL?",
           color="white", size=2.8) +
  annotate("segment", x=4, xend=2.2, y=5.4, yend=4.5,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  annotate("text", x=2.2, y=5.1,
           label="PASS\ncontent_source = NA",
           color="#59A14F", fontface="bold", size=2.8) +
  annotate("segment", x=4, xend=5.8, y=5.4, yend=4.5,
           arrow=arrow(length=unit(3,"mm"), type="closed"),
           color="grey50") +
  annotate("text", x=5.8, y=5.1,
           label="FAIL\ncontent_source \u2260 NA",
           color="#E15759", fontface="bold", size=2.8) +
  annotate("rect", xmin=0.5, xmax=3.8, ymin=3.4, ymax=4.5,
           fill="#59A14F", color="white") +
  annotate("text", x=2.15, y=3.95,
           label="saidit_post\ncontent_source = NA\nfile = filename",
           color="white", size=2.8) +
  annotate("rect", xmin=4.2, xmax=7.5, ymin=3.4, ymax=4.5,
           fill="#E15759", color="white") +
  annotate("text", x=5.85, y=3.95,
           label="BLOCK + ALERT\nPost prevented\nAnomaly flagged for review",
           color="white", size=2.8) +
  coord_cartesian(xlim=c(0, 8), ylim=c(0.5, 11)) +
  theme_void()

left + right +
  plot_annotation(
    title    = "Figure 9.1: Intervention at the Validation Gate",
    subtitle = "Left: current saidit_post_check does not evaluate content_source — anomalous posts pass unchecked\nRight: proposed additional condition blocks any post where content_source is not null",
    theme    = theme(
      plot.title    = element_text(face="bold", size=13),
      plot.subtitle = element_text(size=9, colour="grey50")
    )
  )
```

::: callout-note
## Observations — Intervention at the Validation Gate

- **Figure 9.1 illustrates the specific gap in the current validation gate and the proposed fix**. In the current state (left panel), the `saidit_post_check` evaluates only whether a post can be made - it does not inspect how the content arrived. A post carrying `content_source = HiddenOrca.txt` and `file = NA` passes the check unconditionally, proceeds to publication one second later, and triggers immediate evidence destruction at `+2s` and `+3s`.

- **The proposed intervention (right panel) adds a single condition to the existing gate**. If `content_source` is not null at the point of `saidit_post_check`, the post is blocked and an alert is raised. No new infrastructure is required - the gate already fires at the right moment. The addition is a single null-check on a field that is always `NA` in legitimate posting operations, carrying no risk of false positives.

- Had this condition been active across all three incidents, all three anomalous posts would have been blocked at `+0s` - before publication, before deletion, and before the content reached the SaidIT platform. The one-second window between the check and the post is precisely where the intervention takes effect.
:::

## 10. Conclusion

This investigation applied visual analytics methods to TenantThread's Multi-agent event log to reconstruct three anomalous `saidit_post` events and determine whether they represented a coordinated operation. The findings across Section 5 through Section 9 are summarised below.

**What Was Found**

The three anomalous posts - `HiddenOrca.txt` (May 10), `MellowOtter.txt` (May 11), and `SwiftWren.txt` (May 17) - were not isolated system errors. They were the repeated execution of a single coordinated operation, confirmed by four converging lines of evidence: (1) a stable core actor network present across all three data, (2) a fixed terminal executor (`john_windward`) in every incident, (3) an identical `content_source` delivery mechanism bypassing normal file registration, and (4) a scripted four-event terminal sequnce executing at one-second precision each time.

**Key Actors by Role**

- **Terminal Executor**: `john_windward`. Present in all three incidents as the agent who published each anomalous post.

- **Immediate Delegators**: `chloe_ballast` (May 10 and May 17) and `liam_anchor` (May 11).

- **Stable Core Network**: `victoria_rigging`, `mia_fender`, `lily_anchorline`, `daniel_gangway`, `chloe_ballast`.

- **Earliest Traceable Actor (May 17)**: `james_stern` at `00:51:43 UTC`. The chain limit of the available evidence, not a confirmed point of origin.

**What Remains Unresolved**

- The true originator of the operation cannot be identified from the available data. The relay structure was deliberately designed to prevent back-tracing to a single source.

- The content of the three posted files (what they said and what they meant) cannot be determined. The filenames appear nowhere in `mc2_df` outside the `content_source` field, and no file registration, read, or creation event references them anywhere in the log.

- The sustained virus activity attrributed to `zoey_drydock` beginning May 10 at `20:10:04 UTC` remains uninvestigated. Its relationshop to the posting operation is noted but outside the scope of this study.

**The Intervention**

A single additional null-check on the `content_source` field within the existing `saidit_post_check` gate is sufficient to block all three anomalous posts at the point of validation - before publication, before deletion, and before evidence is destroyed. This is the minimum effective intervention identified by the investigation.
